---
title: "AssignmentMP2"
author: "Srishti Patil"
date: "2024-10-23"
output: html_document
---

```{r}
library(readr)
library(fpp)
library(fpp2)
library(forecast)
```

#Loading dataset 
```{r}
insurance <- read_csv("C:/Users/Srishti/Desktop/FALL24/BF/AssignmentMP2/insurance.csv")
head(insurance)
```
1. Create Timeseries

```{r}
timeSeriesData <- ts(insurance$Quotes, frequency = 12 , start = c(2002,1), end = c(2009,4))
plot(timeSeriesData)
```

2. Verify how much history to include in your analysis. 

As the data shows irregular fluctuations over the time, we will include the entire dataset for better forecasting and analysis.


3. Hypothesize if the dataset has trend, seasonality, or both. 
This time series, spanning approximately 8 years (2002–2009), provides enough history for analysis. The data exhibits regular peaks and troughs, suggesting a strong seasonal component. However, there’s no clear upward or downward trend, indicating that the trend component is likely weak or absent.


4. Verify using Acf
```{r}
Acf(timeSeriesData)
```
The ACF plot shows significant spikes at regular intervals, especially around lags 12 and 24, suggesting an annual seasonal pattern.
The rapid decline of correlation after these lags suggests that there is no strong long-term trend, as a trend would typically result in slower decay in autocorrelation.


5. Verify using decomposition
```{r}
decomposed_data <- decompose(timeSeriesData)
plot(decomposed_data)
```
The decomposition plot also confirms seasonality, as seen in the periodic fluctuations in the "seasonal" component.
The decomposition plot shows a gradual increase and decrease in the trend component over time, indicating a weak trend. The trend is not strong or consistent, but it does fluctuate slightly across the years.



6. Chose an accuracy measure
For the given forecast I choose MAE or MAPE as it would be the best.


7. Create a forecast model for the next 12 months. Include Naive, Average, Exponential Smoothing, HoltWinters, and Decomposition (both types). 

```{r}
# Naive forecast
naive_forecast <- naive(timeSeriesData, h=12)
plot(naive_forecast)

# Mean/average method
mean_forecast <- meanf(timeSeriesData, h=12)
plot(mean_forecast)

# Exponential Smoothing
ses_forecast <- ses(timeSeriesData, h=12)
plot(ses_forecast)


# Holt-Winters forecast
hw_forecast <- HoltWinters(timeSeriesData)
hw_forecast_future <- forecast(hw_forecast, h=12)
plot(hw_forecast_future)

# Moving Averages
MA5_forecast <- ma(timeSeriesData,order=12)
print(MA5_forecast)
plot(MA5_forecast)



# decomposition
decomp_add <- decompose(timeSeriesData, type="additive")
decomp_mult <- decompose(timeSeriesData, type="multiplicative")

# Decomposition Forecasts using trend component
decomp_add_forecast <- forecast(decomp_add$trend, h=12)
decomp_mult_forecast <- forecast(decomp_mult$trend, h=12)

plot(decomp_add_forecast)
plot(decomp_mult_forecast)



mean_accuracy <- accuracy(mean_forecast)
naive_accuracy <- accuracy(naive_forecast)
ses_accuracy <- accuracy(ses_forecast)
hw_accuracy <- accuracy(hw_forecast_future)
decomp_add_accuracy <- accuracy(decomp_add_forecast)
decomp_mult_accuracy <- accuracy(decomp_mult_forecast)



plot(naive_forecast, main="Comparison of Forecast Methods", col="red")
lines(mean_forecast$mean, col="blue")
lines(ses_forecast$mean, col="green")
lines(hw_forecast_future$mean, col="purple")
lines(decomp_add_forecast$mean, col="yellow")
lines(decomp_mult_forecast$mean, col="pink")

```
Due to the trend and seasonality available in the data, MAE and MAPE would generally provide a more stable metric for accuracy. Unlike RMSE, MAE does not penalize larger errors and provides a balanced view of average error. For forecast errors in percentage terms relative to the actual values, MAPE is best forecasting technique. 



8. Show model rank with accuracy measures
Ranking of Models Based on MAE:

Additive Decomposition – 0.1197
Multiplicative Decomposition – 0.1197
Exponential Smoothing – 1.5808
Naive – 1.5989
Holt-Winters – 1.7541
Mean – 1.9239

Hence, The Decomposition (Additive and Multiplicative) models rank the highest and provide the best forecast accuracy based on the given metrics. These models are likely the best choice for forecasting the next 12 months.



```{r}
accuracy_df <- data.frame(
  Model = c( "Additive Decomposition", "Multiplicative Decomposition","Holt-Winters", "Naive", "Mean", "Exponential Smoothing"),
ME= c( decomp_add_accuracy[1], decomp_mult_accuracy[1],hw_accuracy[1], naive_accuracy[1], mean_accuracy[1],ses_accuracy[1]),
  RMSE = c( decomp_add_accuracy[2], decomp_mult_accuracy[2],hw_accuracy[2], naive_accuracy[2], mean_accuracy[2],ses_accuracy[2]),
  MAE = c( decomp_add_accuracy[3], decomp_mult_accuracy[3],hw_accuracy[3], naive_accuracy[3], mean_accuracy[3],ses_accuracy[3]),
  MPE = c( decomp_add_accuracy[4], decomp_mult_accuracy[4],hw_accuracy[4], naive_accuracy[4], mean_accuracy[4],ses_accuracy[4]),
  MAPE = c( decomp_add_accuracy[5], decomp_mult_accuracy[5],hw_accuracy[5], naive_accuracy[5], mean_accuracy[5],ses_accuracy[5]),
  MASE = c( decomp_add_accuracy[6], decomp_mult_accuracy[6],hw_accuracy[6], naive_accuracy[6], mean_accuracy[6],ses_accuracy[6]),
  ACF1 = c( decomp_add_accuracy[7], decomp_mult_accuracy[7],hw_accuracy[7], naive_accuracy[7], mean_accuracy[7],ses_accuracy[7])
)

print(accuracy_df)


```
Given the ranking of models based on accuracy metrics (MAE), I would choose Decomposition(Additive and Multiplicative) or Exponential Smoothing models for forecasting the next 12 months:

###Chosen Models for Forecasting:
Reason:
1.Decomposition (Additive and Multiplicative):
This model performed the best across the key accuracy measures (MAE). The decomposition models capture both the trend and seasonality effectively, which is crucial when your data exhibits these patterns. Decomposition will split the data into the trend, seasonality, and residuals, allowing us to forecast each component separately and combine them to get a more accurate overall forecast.




### How I’m Going to Use These Models for Forecasting:

1. Decomposition Model:
   - First, decompose the historical time series data into its trend, seasonal, and residual components.
   - Forecast each component separately for the next 12 months.
   - Combine the forecasted trend, seasonality, and residual components to generate the final forecast for the next 12 months.

#10. Provide the forecast for the next 12 months (point and range) and explain why you feel confident with these forecasts
```{r}
# Additive decomposition using STL
decomp_add <- stl(timeSeriesData, s.window = "periodic")
decomp_add_forecast <- forecast(decomp_add, h = 12)
print(decomp_add_forecast)
plot(decomp_add_forecast)


# Multiplicative decomposition using STL
log_flight_ts <- log(timeSeriesData)
decomp_mult <- stl(log_flight_ts, s.window = "periodic")
decomp_mult_forecast <- forecast(decomp_mult, h = 12)
decomp_mult_forecast$mean <- exp(decomp_mult_forecast$mean)
print(decomp_mult_forecast)
plot(decomp_mult_forecast)



```

